---
title: "Assimilative causal inference with kalmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assimilative causal inference with kalmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~45 s under rmarkdown::render(), dominated by seven
     causal_information_rate()/aci() calls each quadrating over the full
     lag grid; macOS arm64 (Darwin aarch64), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 150)
set.seed(1)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kalmix)
```

## Why

An agronomist watching a crop-water-stress index through the season, and
suspecting it responds to a slow-moving climate driver, asks a question that
is subtly different from an ordinary correlation: does the future of the
index actually carry information about a hidden climate state, how far
ahead could that give a decision-relevant warning, and can the answer be
trusted, or does the read-out itself need to say when it does not know?
`aci()` answers this by reading a divergence off the two passes `kalmix`
already provides: the Kalman filter and the Rauch-Tung-Striebel smoother of
*Getting started: state-space models with kalmix*. Assimilative causal
inference (Andreou, Chen and Bollt, 2026) casts causal inference as a
Bayesian data-assimilation inverse problem: a hidden state is a cause of an
observed effect at time $t$ exactly when incorporating the *future* of the
effect sharpens the estimate of the state at $t$.

## What

[`aci()`](../reference/aci.html) runs the filter and smoother over an
[`ssm`](../reference/ssm.html) and reads the causal information series and
decision lead-time off the two passes, returning an
[`aci_fit`](../reference/aci_fit.html).
[`causal_information_rate()`](../reference/causal_information_rate.html)
computes the lead-time alone, and is what `aci()` calls internally.
[`innovation_diagnostics()`](../reference/innovation_diagnostics.html) is
the model-adequacy battery `aci()` grounds its verdict through: a Ljung-Box
whiteness test, a scale test on the total normalised innovation, and a
Jarque-Bera normality test, combined conservatively.

## Do

### Causality as an assimilation problem

The filter estimates the state from the past alone, giving
$N(x_{t\mid t}, P_{t\mid t})$; the smoother adds the future, giving
$N(x_{t\mid n}, P_{t\mid n})$. The causal information the measurement
carries about the state at $t$ is the relative entropy (Kullback-Leibler
divergence) of the smoother from the filter,

$$I_t = D_{\mathrm{KL}}\!\left(
  N(x_{t\mid n}, P_{t\mid n}) \,\Vert\, N(x_{t\mid t}, P_{t\mid t}) \right).$$

For two Gaussians this is a closed form, which `kalmix` computes natively.
Time averaged, and integrated over an expanding future window, the
divergence yields a threshold-free *causal information rate*, a decision
lead-time in the time units of the series.

### A known-truth test: a recharge oscillator

The cleanest way to trust a causal read-out is to run it on a system whose
truth is controlled. We simulate a synthetic series from a noise-sustained
recharge oscillator (Jin, 1997), the canonical conceptual model of the El
Nino-Southern Oscillation (ENSO). The hidden state $(T, h)$ is a
sea-surface-temperature anomaly $T$ and a thermocline-depth anomaly $h$;
their coupled, weakly damped dynamics produce an irregular oscillation of
roughly four years. We observe only a scalar that warm $T$ drives through a
coupling `kappa`, standing in for a crop-water-stress index that rises in an
El Nino year. When `kappa` is zero the observation tells us nothing about
the hidden cause.

```{r simulate}
simulate_enso <- function(n = 600, kappa = 0.6, dt = 1 / 12, obs_sd = 0.5) {
  drift <- matrix(c(0, 1.6, -1.6, -0.2), nrow = 2, byrow = TRUE)
  amat <- diag(2) + drift * dt
  z <- matrix(0, nrow = n, ncol = 2)
  x <- numeric(n)
  for (t in 2:n) {
    z[t, ] <- as.numeric(amat %*% z[t - 1, ]) +
      c(rnorm(1, sd = 0.35 * sqrt(dt)), rnorm(1, sd = 0.25 * sqrt(dt)))
    x[t] <- kappa * z[t, 1] + rnorm(1, sd = obs_sd)
  }
  list(x = x, z = z, dt = dt, amat = amat)
}

sim <- simulate_enso(kappa = 0.6)
```

We give `aci()` a state-space model matching the data-generating process. The
observation operator maps the first hidden component (the SST anomaly) to
the observed series through the coupling:

```{r model}
model <- ssm(
  transition = sim$amat,
  observation = matrix(c(0.6, 0), nrow = 1),
  state_cov = diag(c(0.35^2 * sim$dt, 0.25^2 * sim$dt)),
  obs_cov = 0.5^2,
  init_state = c(0, 0),
  init_cov = diag(c(1, 1))
)

fit <- aci(model, sim$x, dt = sim$dt)
fit

recovery_cor <- cor(fit@smoother@smoothed_mean[, 1], sim$z[, 1])
```

Two things to check against the known truth. First, the smoother recovers
the hidden SST anomaly from the observed series *alone*, the inverse-problem
essence, at a correlation of `r round(recovery_cor, 3)` with the simulated
truth. Second, the causal information series should be positive where the
future of the series is informative about the hidden state.

```{r info-plot, echo = has_ggplot2, eval = has_ggplot2, fig.cap = "Top: the hidden ENSO state recovered from the observed index alone, against the truth used to simulate it. Bottom: the per-step causal information series that recovery is built from."}
years <- (seq_along(sim$x) - 1) * sim$dt
plot_data <- rbind(
  data.frame(year = years, value = sim$z[, 1], series = "truth (simulated)",
             panel = "Hidden ENSO state"),
  data.frame(year = years, value = fit@smoother@smoothed_mean[, 1],
             series = "smoother estimate", panel = "Hidden ENSO state"),
  data.frame(year = years, value = fit@causal_information,
             series = "causal information",
             panel = "Causal information I_t (nats)")
)
panel_levels <- c("Hidden ENSO state", "Causal information I_t (nats)")
plot_data$panel <- factor(plot_data$panel, levels = panel_levels)

ggplot2::ggplot(plot_data, ggplot2::aes(x = year, y = value, colour = series)) +
  ggplot2::geom_line(linewidth = 0.5) +
  ggplot2::facet_wrap(~panel, ncol = 1, scales = "free_y") +
  ggplot2::scale_colour_viridis_d(name = NULL, end = 0.8) +
  ggplot2::labs(
    x = "year", y = NULL,
    title = "Recovered state and causal information, coupled system"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

```{r info-plot-skip, echo = FALSE, eval = !has_ggplot2, results = "asis"}
cat("ggplot2 is not installed on this build, so the recovery-and-information",
    "figure is skipped.\n")
```

The smoother trace tracks the truth closely in the top panel although the
model never observes the state directly, and the causal-information trace
in the bottom panel rises and falls with the oscillation rather than
sitting flat at zero, which is the discriminating signature the null case
below is checked against.

### The null carries no causal information

Under the null of no coupling between the hidden state and the observation,
the future of the series adds nothing the filter did not already have, so
the causal information collapses to zero and the lead-time with it:

```{r null}
sim_null <- simulate_enso(kappa = 0)
model_null <- ssm(
  transition = sim_null$amat,
  observation = matrix(c(1e-6, 0), nrow = 1),
  state_cov = diag(c(0.35^2 * sim_null$dt, 0.25^2 * sim_null$dt)),
  obs_cov = 0.5^2,
  init_state = c(0, 0),
  init_cov = diag(c(1, 1))
)
fit_null <- aci(model_null, sim_null$x, dt = sim_null$dt)
```

```{r null-kable, echo = FALSE}
knitr::kable(
  data.frame(
    system = c("coupled (kappa = 0.6)", "null (kappa = 0)"),
    mean_causal_information = round(
      c(fit@mean_causal_information, fit_null@mean_causal_information), 4
    ),
    lead_time_years = round(c(fit@lead_time, fit_null@lead_time), 3)
  ),
  col.names = c("System", "Mean causal information (nats)",
                "Lead-time (years)"),
  caption = paste(
    "The coupled system carries causal information and a positive",
    "lead-time; the null carries neither."
  )
)
```

### A divergence between two Gaussians, cross-checked

The metric `aci()` reads off the two passes is the relative entropy between
the smoother and filter Gaussians at each step. This is canonical Gaussian
arithmetic, so `kalmix` implements it natively and stays self-contained.
Where the `kernR` package is installed, its `relative_entropy()` provides an
independent check on the value: the package's test suite asserts the two
agree and skips the comparison when `kernR` is absent, so `kalmix` builds
and checks cleanly either way.

### Grounding the verdict: abstention under model error

ACI reads its causal verdict off a state-space model, so the verdict is only
as good as the model. A model that cannot even describe the observed series
cannot support a causal claim about it. `aci()` therefore grounds every
read-out through `innovation_diagnostics()`, run on the standardised
one-step-ahead innovations. When the innovations reject the model it is
declared inadequate and the verdict abstains, with a distinct grounding
token instead of a confident lead-time:

```{r abstain}
## A frozen-state local level cannot track the oscillating ENSO index: its
## innovations are autocorrelated and mis-scaled, so ACI abstains.
frozen <- ssm(transition = 1, observation = 1, state_cov = 1e-6, obs_cov = 1,
              init_state = 0, init_cov = 10)
abstained <- aci(frozen, sim$x, dt = sim$dt)
abstained@grounding
abstained@grounding_reason
abstained@adequacy@adequate
abstention_classes <- class(abstained)
abstention_classes
```

```{r grounded}
## Adequate but unverified by default; grounded only with dated provenance.
adequate_unverified <- aci(model, sim$x, dt = sim$dt)@grounding
adequate_grounded <- aci(
  model, sim$x, dt = sim$dt,
  mechanism = list(source = "validated against the OSSE truth",
                    verified_on = as.Date("2026-06-14"))
)@grounding
```

### The lead-time without re-smoothing

The decision lead-time integrates a divergence profile over an expanding
future window. Formed directly, each window length reruns the smoother on a
truncated series. `kalmix` instead grows the smoothed estimate of the
anchor state forward in place as each later observation is assimilated (a
fixed-point smoother recursion that accumulates the Rauch-Tung-Striebel
gains), so the whole profile costs one filter-smoother pass and a light
forward recursion per anchor, with no re-smoothing. This `online` engine is
the default; the `expanding` engine is retained as an independent
cross-check:

```{r engines}
cir_online <- causal_information_rate(model, sim$x, dt = sim$dt,
                                       engine = "online")
cir_expanding <- causal_information_rate(model, sim$x, dt = sim$dt,
                                          engine = "expanding")
```

```{r engines-kable, echo = FALSE}
knitr::kable(
  data.frame(
    engine = c("online (default)", "expanding"),
    lead_time_years = round(c(cir_online, cir_expanding), 6)
  ),
  col.names = c("Engine", "Lead-time (years)"),
  caption = paste(
    "The two engines agree to numerical precision on the coupled system."
  )
)
```

Because the online recursion never truncates the series, it also applies to
time-varying state-space models, where the expanding construction cannot: a
model whose system matrices change step by step still yields a lead-time
from a single forward pass. The two engines agreeing shows only that the
smoother recursion is implemented consistently between them: they share the
quadrature and the reduction, so their agreement is not evidence about the
range functional itself. That functional is graded separately, in the
package's own conformance suite (`tests/testthat/test-cir-conformance.R`),
which reduces `acir`'s own divergence profile with `kalmix`'s reducer and
requires the objective and subjective ranges of Andreou, Chen and Bollt
(2026) to come back; `acir` is in turn graded cell-by-cell against the
method authors' MATLAB reference. That chain, back to the authors' own
code, is what makes it an oracle here, and it is not recomputed in this
vignette.

### A real ENSO index

The known-truth oscillator proves the read-out is internally consistent; the
package's test suite also runs `aci()` on the real NOAA Oceanic Nino Index
(ONI) as an external check, in `tests/testthat/test-aci.R`, on a fixture
shipped with the package (`tests/testthat/fixtures/oni_noaa.csv`, the CPC
ONI series, 1980 to 2020, public-domain US Government data, NOAA Climate
Prediction Center). The same model and data are reproduced here directly,
rather than cited unrun, so the number below is computed in this vignette:

```{r oni-load}
oni_path <- file.path("..", "tests", "testthat", "fixtures", "oni_noaa.csv")
has_oni <- file.exists(oni_path)
```

```{r oni-fit, eval = has_oni, echo = has_oni}
oni <- read.csv(oni_path, stringsAsFactors = FALSE)$oni

## A two-state local oscillator: a damped rotation captures the quasi-
## periodic ENSO swing, observed through the index with measurement noise.
dt_year <- 1 / 12
angle <- 2 * pi * dt_year / 3.8
rho <- 0.96
rot <- rho * matrix(
  c(cos(angle), -sin(angle), sin(angle), cos(angle)),
  nrow = 2L, byrow = TRUE
)
oni_model <- ssm(
  transition = rot,
  observation = matrix(c(1, 0), nrow = 1L),
  state_cov = diag(c(0.05, 0.05)),
  obs_cov = 0.2,
  init_state = c(oni[1L], 0),
  init_cov = diag(c(1, 1))
)
oni_fit <- aci(oni_model, oni, dt = dt_year)
oni_tracked <- cor(oni_fit@smoother@smoothed_mean[, 1L], oni)
```

```{r oni-skip, eval = !has_oni, echo = FALSE, results = "asis"}
cat("The NOAA ONI fixture is not reachable at the path this vignette",
    "expects, so the real-index read-out is skipped; the model and result",
    "are recorded in `tests/testthat/test-aci.R` instead.\n")
```

```{r oni-kable, echo = FALSE, eval = has_oni}
knitr::kable(
  data.frame(
    quantity = c("smoother-to-index correlation", "decision lead-time (years)"),
    value = round(c(oni_tracked, oni_fit@lead_time), 3)
  ),
  col.names = c("Quantity", "Value"),
  caption = "ACI on the real NOAA ONI series, 1980 to 2020."
)
```

Two things bound how far that figure should be pushed. The quadrature is
evaluated at every lag in the window rather than at a dozen points, and
warns when a grid the caller thins has not converged, so the number is not
grid-dependent at the default. And an anchor whose future window runs past
the end of the record is right-censored: its range is integrated over the
lags the record supports and returned as a lower bound with `censored` set,
rather than extrapolated. A rolling read-out near the end of a record is
censored by construction, and a censored lead-time cannot carry a grounded
verdict.

## Read

```{r read-prep, echo = FALSE}
oni_sentence <- if (has_oni) {
  sprintf(
    paste(
      "On the real NOAA ONI series the smoother tracks the observed",
      "index at a correlation of %s, and the decision lead-time comes",
      "out at %s years, the kind of sub-decadal predictability horizon",
      "the ENSO literature reports."
    ),
    round(oni_tracked, 3), round(oni_fit@lead_time, 3)
  )
} else {
  paste(
    "The real-index section above explains why that read-out was",
    "skipped on this build."
  )
}
```

On the simulated oscillator, the smoother recovers the hidden state at a
correlation of `r round(recovery_cor, 3)` with the truth it never observes
directly. The coupled system carries a mean causal information of
`r round(fit@mean_causal_information, 4)` nats and a lead-time of
`r round(fit@lead_time, 3)` years, while the null carries
`r round(fit_null@mean_causal_information, 4)` nats and a lead-time of
`r round(fit_null@lead_time, 3)`: the coupling is what the causal-information
read-out is discriminating, not an artefact of the recursion.

The abstention example carries `abstained@grounding` equal to
`r sprintf("\x60%s\x60", abstained@grounding)`, with reason
`r sprintf("\x60%s\x60", abstained@grounding_reason)`, because the frozen
local level's innovations fail `innovation_diagnostics()`
(`abstained@adequacy@adequate` is `r abstained@adequacy@adequate`). The
returned object carries the classes
`r paste(sprintf("\x60%s\x60", abstention_classes), collapse = ", ")`,
so a caller elsewhere in the orchestra can recognise this as a structured
refusal, `kalmix_abstention`, without parsing the grounding string. Passing
the diagnostics is a different matter from having a verified mechanism: the
same coupled model with no declared provenance carries grounding
`r sprintf("\x60%s\x60", adequate_unverified)`, and only rises to
`r sprintf("\x60%s\x60", adequate_grounded)` once a dated mechanism record
is supplied. An adequate model is not thereby a verified one; the two
chunks above show both halves of that distinction on the same model.

The online and expanding engines agree on the coupled system to within
`r formatC(abs(cir_online - cir_expanding), format = "e", digits = 1)`
years, floating-point precision rather than a genuine numerical difference,
confirming they implement the same smoother recursion. `r oni_sentence`

## Limits

Every read-out here is only as trustworthy as the state-space model behind
it, and `aci()`'s adequacy check establishes self-consistency, not
correspondence with the world: an adequate model that was never checked
against an external authority stays labelled as unverified rather than
grounded, exactly as shown above. The oscillator example is a known-truth
system built to demonstrate the read-out, not a validated ENSO model, and
the real-index section, while run on genuine public data, uses a
deliberately simple damped-rotation model rather than a dynamical model
fitted to the physical system. Under a Student-t observation family the
filter pass is the outlier-robust variational update, whose per-step
posteriors are Gaussian approximations; the causal information and
lead-time then inherit that approximation rather than being exact. A
regime-switching model compares the collapsed Gaussian posteriors of the
Kim smoother and the GPB1 filter rather than the true mixtures, and its
lead-time uses the expanding-window construction only, because the online
recursion is linear-Gaussian only. The online and expanding engines
agreeing on this vignette's models is evidence the smoother recursion is
implemented consistently between them, not evidence that the underlying
range functional matches the source paper; that grading is external, cited
above and not recomputed here.

## What to read next

*Getting started: state-space models with kalmix* covers the
`kalman_filter()` and `rts_smoother()` passes this vignette reads a
divergence between. *Change-points and hidden states with kalmix* and
*Pairs trading with kalmix* cover the discrete-state and regime-switching
models an `aci()` call can also be given. *The Elliott-van der Hoek-Malcolm
spread model with kalmix* shows the same `ssm_fit()` machinery this
vignette's models were specified by hand instead of estimating.

## Reproduce

Seed `1` for the simulated oscillator; the real-index section draws no
random numbers.

```{r session-info}
sessionInfo()
```
