---
title: "Assimilative causal inference with kalmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assimilative causal inference with kalmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
set.seed(1)
library(kalmix)
```

The interrupted-time-series contrast of `its_causal()` asks what an
intervention *did*. Assimilative causal inference asks a different question:
across a continuously observed series, *when* and *how far ahead* does a hidden
driver carry causal information about what we measure. `aci()` answers it by
reading a divergence off the two passes `kalmix` already provides -- the Kalman
filter and the Rauch-Tung-Striebel smoother.

## Causality as an assimilation problem

Andreou, Chen and Bollt (2026) cast causal inference as a Bayesian
data-assimilation inverse problem. A hidden state is a cause of an observed
effect at time $t$ exactly when incorporating the *future* of the effect
sharpens our estimate of the state at $t$. The filter estimates the state from
the past alone, giving $N(x_{t\mid t}, P_{t\mid t})$; the smoother adds the
future, giving $N(x_{t\mid n}, P_{t\mid n})$. The causal information the
measurement carries about the state at $t$ is the relative entropy
(Kullback-Leibler divergence) of the smoother from the filter,

$$I_t = D_{\mathrm{KL}}\!\left(
  N(x_{t\mid n}, P_{t\mid n}) \,\Vert\, N(x_{t\mid t}, P_{t\mid t}) \right).$$

For two Gaussians this is a closed form, which `kalmix` computes natively. Time
averaged, and integrated over an expanding future window, the divergence yields
a threshold-free *causal information rate* -- a decision lead-time in the time
units of the series.

## A known-truth test: a recharge oscillator

The cleanest way to trust a causal read-out is to run it on a system whose truth
we control. We use a noise-sustained recharge oscillator (Jin, 1997), the
canonical conceptual model of the El Nino-Southern Oscillation (ENSO). The
hidden state $(T, h)$ is a sea-surface-temperature anomaly $T$ and a
thermocline-depth anomaly $h$; their coupled, weakly damped dynamics produce an
irregular oscillation of roughly four years. We observe only a scalar that warm
$T$ drives through a coupling `kappa` -- think of a crop-water-stress index that
rises in an El Nino year. When `kappa` is zero the observation tells us nothing
about the hidden cause.

```{r simulate}
simulate_enso <- function(n = 600, kappa = 0.6, dt = 1 / 12, obs_sd = 0.5) {
  drift <- matrix(c(0, 1.6, -1.6, -0.2), nrow = 2, byrow = TRUE)
  amat <- diag(2) + drift * dt
  z <- matrix(0, nrow = n, ncol = 2)
  x <- numeric(n)
  for (t in 2:n) {
    z[t, ] <- as.numeric(amat %*% z[t - 1, ]) +
      c(rnorm(1, sd = 0.35 * sqrt(dt)), rnorm(1, sd = 0.25 * sqrt(dt)))
    x[t] <- kappa * z[t, 1] + rnorm(1, sd = obs_sd)
  }
  list(x = x, z = z, dt = dt, amat = amat)
}

sim <- simulate_enso(kappa = 0.6)
```

We give `aci()` a state-space model matching the data-generating process. The
observation operator maps the first hidden component (the SST anomaly) to the
observed series through the coupling:

```{r model}
model <- ssm(
  transition = sim$amat,
  observation = matrix(c(0.6, 0), nrow = 1),
  state_cov = diag(c(0.35^2 * sim$dt, 0.25^2 * sim$dt)),
  obs_cov = 0.5^2,
  init_state = c(0, 0),
  init_cov = diag(c(1, 1))
)

fit <- aci(model, sim$x, dt = sim$dt)
fit
```

Two things to check against the known truth. First, the smoother recovers the
hidden SST anomaly from the observed series *alone* -- the inverse-problem
essence:

```{r recovery}
cor(fit@smoother@smoothed_mean[, 1], sim$z[, 1])
```

Second, the causal information series is positive where the future of the
series is informative about the hidden state, and the decision lead-time is a
sensible, sub-decadal horizon:

```{r info-plot, echo = FALSE}
years <- (seq_along(sim$x) - 1) * sim$dt
op <- par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
plot(years, sim$z[, 1], type = "l", col = "navy", xlab = "", ylab = "ENSO state T",
     main = "Hidden ENSO state (navy) vs smoother estimate from the index alone (orange)")
lines(years, fit@smoother@smoothed_mean[, 1], col = "orange")
abline(h = 0, lty = 3, col = "grey60")
plot(years, fit@causal_information, type = "l", col = "firebrick",
     xlab = "year", ylab = "causal information (nats)",
     main = "ACI causal information series I_t = KL(smoother || filter)")
par(op)
```

## The null carries no causal information

Under the null -- no coupling between the hidden state and the observation --
the future of the series adds nothing the filter did not already have, so the
causal information collapses to zero and the lead-time with it. This is the
discriminating behaviour a causal read-out must show:

```{r null}
sim_null <- simulate_enso(kappa = 0)
model_null <- ssm(
  transition = sim_null$amat,
  observation = matrix(c(1e-6, 0), nrow = 1),
  state_cov = diag(c(0.35^2 * sim_null$dt, 0.25^2 * sim_null$dt)),
  obs_cov = 0.5^2,
  init_state = c(0, 0),
  init_cov = diag(c(1, 1))
)
fit_null <- aci(model_null, sim_null$x, dt = sim_null$dt)
c(coupled = fit@mean_causal_information, null = fit_null@mean_causal_information)
c(coupled_lead = fit@lead_time, null_lead = fit_null@lead_time)
```

The coupled system carries causal information and a positive lead-time; the null
carries neither.

## A divergence between two Gaussians, cross-checked

The metric `aci()` reads off the two passes is the relative entropy between the
smoother and filter Gaussians at each step. This is canonical Gaussian
arithmetic, so `kalmix` implements it natively and stays self-contained. Where
the orchestra's `kernR` package is installed, its `relative_entropy()` provides
an independent check on the value -- the package's test suite asserts the two
agree and skips the comparison when `kernR` is absent, so `kalmix` builds and
checks cleanly either way.

## Grounding the verdict: abstention under model error

ACI reads its causal verdict off a state-space model, so the verdict is only as
honest as the model. A model that cannot even describe the observed series
cannot support a causal claim about it. `aci()` therefore grounds every read-out
through `innovation_diagnostics()`, the standard battery of state-space model
checks run on the standardised one-step-ahead innovations: a Ljung-Box test for
serial correlation, a $\chi^2$ test of the total normalised innovation squared
for a mis-scaled noise level, and a Jarque-Bera test for non-Gaussianity. When
the innovations reject the model it is declared inadequate and the verdict
abstains with a `[unverified]` grounding token instead of asserting a confident
lead-time.

```{r abstain}
# A frozen-state local level cannot track the oscillating ENSO index: its
# innovations are autocorrelated and mis-scaled, so ACI abstains.
frozen <- ssm(transition = 1, observation = 1, state_cov = 1e-6, obs_cov = 1,
              init_state = 0, init_cov = 10)
abstained <- aci(frozen, sim$x, dt = sim$dt)
c(grounding = abstained@grounding, reason = abstained@grounding_reason)
abstained@adequacy
```

Passing the diagnostics establishes self-consistency only -- it does not verify
the model against the outside world, which is what grounding means. An adequate
model is labelled `"grounded"` solely when it is accompanied by declared, dated
mechanism provenance: a record that the model was checked against the external
authority that owns it. Without that, the honest label stays `[unverified]`,
even when the model fits.

```{r grounded}
# Adequate but unverified by default; grounded only with dated provenance.
aci(model, sim$x, dt = sim$dt)@grounding
aci(model, sim$x, dt = sim$dt,
    mechanism = list(source = "validated against the OSSE truth",
                     verified_on = as.Date("2026-06-14")))@grounding
```

This is ACI's guard against its own central limitation: the verdict is only as
good as the model, so a model the data reject -- or one whose standing as the
true mechanism has never been checked -- is never allowed to pass as a grounded
causal claim.

## The lead-time without re-smoothing

The decision lead-time integrates a divergence profile over an expanding future
window. Formed directly, each window length reruns the smoother on a truncated
series. `kalmix` instead grows the smoothed estimate of the anchor state forward
in place as each later observation is assimilated -- a fixed-point smoother
recursion that accumulates the Rauch-Tung-Striebel gains -- so the whole profile
costs one filter-smoother pass and a light forward recursion per anchor, with no
re-smoothing. This `online` engine is the default; the `expanding` engine is
retained as an independent cross-check, and the two agree to numerical
precision.

```{r engines}
c(online    = causal_information_rate(model, sim$x, dt = sim$dt,
                                      engine = "online"),
  expanding = causal_information_rate(model, sim$x, dt = sim$dt,
                                      engine = "expanding"))
```

Because the online recursion never truncates the series, it also applies to
time-varying state-space models, where the expanding construction cannot: a
model whose system matrices change step by step still yields a lead-time from a
single forward pass.

## A real ENSO index

The known-truth oscillator proves the read-out is internally consistent. As an
external check, the package's test suite also runs `aci()` on the real NOAA
Oceanic Nino Index -- a recorded copy ships as a test fixture, and an
online-gated test confirms the read-out behaves on data nobody in this package
generated. On the recorded index the smoother tracks the observed ENSO swing
closely and the decision lead-time lands near half a year, the kind of
predictability horizon the ENSO literature reports.

## References

Andreou, M., Chen, N. and Bollt, E. (2026). Assimilative causal inference.
*Nature Communications*, 17, 1854.

Jin, F.-F. (1997). An equatorial ocean recharge paradigm for ENSO. Part I: A
conceptual model. *Journal of the Atmospheric Sciences*, 54(7), 811--829.

Kalman, R. E. (1960). A new approach to linear filtering and prediction
problems. *Journal of Basic Engineering*, 82(1), 35--45.

Rauch, H. E., Tung, F. and Striebel, C. T. (1965). Maximum likelihood estimates
of linear dynamic systems. *AIAA Journal*, 3(8), 1445--1450.
