---
title: "Change-points and hidden states with kalmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Change-points and hidden states with kalmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~1 s under rmarkdown::render(); macOS arm64 (Darwin
     aarch64), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 150)
set.seed(1)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kalmix)
```

## Why

A paddock's canopy-vigour index climbs across the season and then something
changes: did the index really shift to a new level partway through, at an
identifiable point in time, or is that just the ordinary wobble a flat
series always produces on its own? And separately, if a pest-pressure index
instead flips back and forth between a quiet background level and a
flare-up, is that better read as a recurring pair of states than as a
one-off shift? `kalmix` answers the first question with change-point
detection and the second with a hidden Markov model; both are shown here on
synthetic series built to isolate the mechanism. The regime-switching
mixture filter, a third flavour of state that switches between several
linear-Gaussian dynamics rather than two fixed emission levels, is shown in
*Pairs trading with kalmix*.

## What

[`detect_changepoint()`](../reference/detect_changepoint.html) finds shifts
in the mean of a sequence by binary segmentation on a likelihood-ratio
statistic, returning a [`changepoint_fit`](../reference/changepoint_fit.html)
of the split locations and the fitted segment means.
[`hmm()`](../reference/hmm.html) builds a finite Markov chain over hidden
states, each emitting a Gaussian observation.
[`hmm_filter()`](../reference/hmm_filter.html) gives the smoothed state
posteriors and the model log-likelihood, and
[`hmm_viterbi()`](../reference/hmm_viterbi.html) returns the single most
likely state path, jointly optimal over the whole series rather than
step-by-step.

## Do

### Change-point detection

At split $\tau$ of a length-$n$ segment, `detect_changepoint()`'s statistic
is the drop in the Gaussian twice-negative-profile-log-likelihood between a
one-mean and a two-mean fit,

$$n \log \hat\sigma^2_{1:n} - \tau \log \hat\sigma^2_{1:\tau} -
  (n - \tau) \log \hat\sigma^2_{\tau+1:n},$$

with the variance estimated within each candidate segment; the split
location that maximises this statistic is the candidate change-point, kept
only when the statistic clears a penalty (Scott and Knott, 1974). We
simulate a canopy-vigour index with two known shifts, a lift at index 50 and
a partial decline at index 100, and recover them:

```{r changepoint}
vigour <- c(rnorm(50, mean = 0), rnorm(50, mean = 4), rnorm(50, mean = 1))
cp <- detect_changepoint(vigour)
cp
```

```{r changepoint-kable, echo = FALSE}
knitr::kable(
  data.frame(
    segment = seq_along(cp@segment_means),
    fitted_mean = round(cp@segment_means, 3)
  ),
  col.names = c("Segment", "Fitted mean"),
  caption = "The fitted mean of each segment `detect_changepoint()` recovered."
)
```

```{r changepoint-plot, echo = has_ggplot2, eval = has_ggplot2, fig.cap = "The simulated canopy-vigour index (grey points), the fitted segment means (solid line) and the detected change-points (dashed lines)."}
bounds <- c(0, cp@changepoints, length(vigour))
seg_data <- data.frame(
  xstart = bounds[-length(bounds)] + 1,
  xend = bounds[-1],
  fitted_mean = cp@segment_means
)
point_data <- data.frame(index = seq_along(vigour), vigour = vigour)
cp_data <- data.frame(index = cp@changepoints + 0.5)

ggplot2::ggplot() +
  ggplot2::geom_point(
    data = point_data, ggplot2::aes(x = index, y = vigour),
    colour = "grey55", size = 0.9
  ) +
  ggplot2::geom_segment(
    data = seg_data,
    ggplot2::aes(x = xstart, xend = xend, y = fitted_mean, yend = fitted_mean),
    colour = "#1B9E77", linewidth = 1
  ) +
  ggplot2::geom_vline(
    data = cp_data, ggplot2::aes(xintercept = index),
    linetype = "dashed", colour = "#7570B3"
  ) +
  ggplot2::labs(
    x = "day of season", y = "canopy-vigour index (synthetic units)",
    title = "Detected change-points in a synthetic vigour index"
  ) +
  ggplot2::theme_minimal()
```

```{r changepoint-plot-skip, echo = FALSE, eval = !has_ggplot2, results = "asis"}
cat("ggplot2 is not installed on this build, so the change-point figure is",
    "skipped.\n")
```

### The penalty controls over-segmentation

The hard part of change-point detection is not finding a split; it is
deciding when *not* to. On a single homogeneous segment the best available
split still improves the fit by chance, so an unpenalised search would
shatter a flat series into spurious pieces. `detect_changepoint()` guards
against this with a penalty a split must clear. The default `"mbic"`
penalty is calibrated so a genuinely flat index is left whole:

```{r changepoint-null}
flat <- rnorm(200)
n_flat_cp <- length(detect_changepoint(flat)@changepoints)
```

A weaker penalty is available for callers who want more sensitivity, at the
cost of more false positives, on a subtler shift than the one simulated
above:

```{r changepoint-penalty}
subtle <- c(rnorm(60, 0), rnorm(60, 1.2), rnorm(60, 0))
n_mbic <- length(detect_changepoint(subtle, penalty = "mbic")@changepoints)
n_sic <- length(detect_changepoint(subtle, penalty = "sic")@changepoints)
```

```{r penalty-kable, echo = FALSE}
knitr::kable(
  data.frame(
    series = c("flat (no shift)", "subtle (1.2-SD shift, twice)"),
    mbic = c(n_flat_cp, n_mbic),
    sic = c(NA_integer_, n_sic)
  ),
  col.names = c("Series", "Change-points (mbic)", "Change-points (sic)"),
  caption = paste(
    "How many change-points each penalty reports on a flat and a",
    "subtly-shifted series."
  )
)
```

### Hidden Markov models

A hidden Markov model is the discrete-state cousin of the state-space model:
a finite Markov chain over hidden states, each emitting a Gaussian
observation. We build a two-state pest-pressure model, a quiet background
state near `0` and a flare-up state near `4`, and simulate a season in which
pressure flares for a central block:

```{r hmm-build}
model <- hmm(
  init_prob = c(0.5, 0.5),
  transition = matrix(c(0.95, 0.05, 0.05, 0.95), nrow = 2, byrow = TRUE),
  emission_mean = c(0, 4), emission_sd = c(1, 1)
)
model

true_state <- c(rep(1L, 60), rep(2L, 60), rep(1L, 60))
y <- rnorm(180, mean = c(0, 4)[true_state], sd = 1)
```

The smoothed posterior and the Viterbi path both recover the hidden pressure
state:

```{r hmm-recover}
fit <- hmm_filter(model, y)
path <- hmm_viterbi(model, y)
acc_smoothed <- mean(fit@state_path == true_state)
acc_viterbi <- mean(path == true_state)
```

```{r hmm-kable, echo = FALSE}
knitr::kable(
  data.frame(
    readout = c("smoothed mode", "Viterbi path"),
    accuracy = round(c(acc_smoothed, acc_viterbi), 3)
  ),
  col.names = c("Read-out", "Accuracy against the true state"),
  caption = paste(
    "Agreement of each state read-out with the simulated true pressure",
    "state."
  )
)
```

```{r hmm-plot, echo = has_ggplot2, eval = has_ggplot2, fig.cap = "Observations (points, coloured by true hidden state) with the smoothed posterior probability of the flare-up state (line, right-hand axis)."}
range_y <- range(y)
span_y <- diff(range_y)
post_high <- fit@smoothed[, 2]
post_scaled <- post_high * span_y + min(range_y)

hmm_data <- data.frame(
  day = seq_along(y), y = y,
  state = factor(c("background", "flare-up")[true_state],
                 levels = c("background", "flare-up")),
  posterior = post_scaled
)

ggplot2::ggplot(hmm_data, ggplot2::aes(x = day)) +
  ggplot2::geom_point(ggplot2::aes(y = y, colour = state), size = 1) +
  ggplot2::geom_line(ggplot2::aes(y = posterior), colour = "black",
                      linewidth = 0.6) +
  ggplot2::scale_colour_viridis_d(name = "true state", end = 0.8) +
  ggplot2::scale_y_continuous(
    name = "pest-pressure index (synthetic units)",
    sec.axis = ggplot2::sec_axis(
      ~ (. - min(range_y)) / span_y, name = "P(flare-up state)"
    )
  ) +
  ggplot2::labs(x = "day of season",
                title = "HMM: smoothed posterior of the flare-up state") +
  ggplot2::theme_minimal()
```

```{r hmm-plot-skip, echo = FALSE, eval = !has_ggplot2, results = "asis"}
cat("ggplot2 is not installed on this build, so the HMM posterior figure is",
    "skipped.\n")
```

## Read

`detect_changepoint()` placed `r length(cp@changepoints)` change-points on
the simulated vigour index, at indices
`r paste(cp@changepoints, collapse = " and ")`, close to the true shift
locations of 50 and 100, and recovered three segment means that step from
low to high and back down again, shown in the table above and visible in
the figure as the solid fitted line jumping at each dashed change-point.
On a genuinely flat series the default `"mbic"` penalty reports
`r n_flat_cp` change-points; on the subtler 1.2-standard-deviation shift the
weaker `"sic"` penalty finds `r n_sic` against `r n_mbic` for `"mbic"`,
which is the direct cost of the extra sensitivity: fewer missed shifts, more
willingness to call one.

Both hidden-Markov-model read-outs recover the true pressure state with
`r sprintf("%.1f", 100 * acc_smoothed)` per cent (smoothed mode) and
`r sprintf("%.1f", 100 * acc_viterbi)` per cent (Viterbi path) accuracy on
these well-separated states. The figure makes the difference between the two
read-outs visible rather than just numeric: the black posterior line rises
towards one only within the flare-up block and returns to near zero either
side of it, tracking the coloured points closely, while the Viterbi path
(not itself plotted, being a hard assignment rather than a probability) is
the single state sequence that is jointly most likely, not merely the state
that is individually most likely at each step.

## Limits

Every series in this vignette is synthetic and built to make its own
mechanism visible: the segments and states are well separated by
construction, which favours both methods more than a real canopy or pest
record generally would. `detect_changepoint()` assumes Gaussian segments of
constant variance; a shift in variance rather than mean, or a genuinely
non-Gaussian series, is outside what the statistic tests. The hidden Markov
model here has two states with dynamics and emission parameters fixed by
the caller rather than estimated from data (`hmm_fit()` covers estimation
and is not exercised in this vignette); a real pest or disease record would
need that fitting step, and its number of states would itself be a modelling
choice this vignette does not address. Neither method here carries the
model-adequacy grounding that `aci()` attaches to its causal read-out; that
machinery, including the abstention it produces on a genuinely inadequate
model, is shown in *Assimilative causal inference with kalmix*.

## What to read next

*Getting started: state-space models with kalmix* covers the continuously
evolving linear-Gaussian state these discrete-state models sit alongside.
*Pairs trading with kalmix* shows the third flavour of discrete structure,
the regime-switching mixture filter, applied to a mean-reverting spread.
*Assimilative causal inference with kalmix* grounds a causal read-out
against a model-adequacy check that can abstain, which neither method here
performs.

## Reproduce

Seed `1` for every simulation in this vignette.

```{r session-info}
sessionInfo()
```
