---
title: "Change-points and hidden states with kalmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Change-points and hidden states with kalmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
set.seed(1)
library(kalmix)
```

Many series are not smooth: they switch abruptly between states or shift
their mean at unknown times. `kalmix` handles three flavours of this:
change-point detection for abrupt shifts in level, hidden Markov models for
recurring discrete states, and the regime-switching mixture filter for series
that switch between several linear-Gaussian dynamics. This vignette walks
through the first two; the mixture filter is shown in *Pairs trading with
kalmix*.

## Change-point detection

`detect_changepoint()` finds shifts in the mean of a sequence by binary
segmentation (Scott and Knott, 1974) on the single-change-point
likelihood-ratio statistic. At split $\tau$ of a length-$n$ segment the
statistic is the drop in the Gaussian twice-negative-profile-log-likelihood
between a one-mean and a two-mean fit,

$$n \log \hat\sigma^2_{1:n} - \tau \log \hat\sigma^2_{1:\tau} -
  (n - \tau) \log \hat\sigma^2_{\tau+1:n},$$

with the variance estimated within each candidate segment; the split
location that maximises this statistic is the candidate change-point, kept
only when the statistic clears a penalty. We simulate a series with two
known shifts, at index 50 and index 100, and recover them:

```{r changepoint}
x <- c(rnorm(50, mean = 0), rnorm(50, mean = 4), rnorm(50, mean = 1))
cp <- detect_changepoint(x)
cp
```

```{r changepoint-plot, echo = FALSE, fig.cap = "Detected change-points (blue dashed lines) against the simulated series, with the fitted segment means (red) overlaid."}
plot(x, col = "grey60", pch = 16, cex = 0.6, xlab = "time", ylab = "value",
     main = "Detected change-points")
bounds <- c(0, cp@changepoints, length(x))
for (i in seq_along(cp@segment_means)) {
  segments(bounds[i] + 1, cp@segment_means[i], bounds[i + 1],
           cp@segment_means[i], col = "firebrick", lwd = 2)
}
abline(v = cp@changepoints + 0.5, lty = 2, col = "steelblue")
```

The detector places its change-points within a couple of indices of the truth
and recovers the three segment means.

### The penalty controls over-segmentation

The hard part of change-point detection is not finding a split. It is
deciding when *not* to. On a single homogeneous segment the best available
split still improves the fit by chance, so an unpenalised search would shatter
a flat series into spurious pieces. `detect_changepoint()` guards against this
with a penalty that a split must clear. The default `"mbic"` penalty is
calibrated so a genuinely homogeneous series is left whole:

```{r changepoint-null}
flat <- rnorm(200)
detect_changepoint(flat)@changepoints
```

A weaker penalty is available for callers who want more sensitivity, at the
cost of more false positives:

```{r changepoint-penalty}
subtle <- c(rnorm(60, 0), rnorm(60, 1.2), rnorm(60, 0))
c(mbic = length(detect_changepoint(subtle, penalty = "mbic")@changepoints),
  sic  = length(detect_changepoint(subtle, penalty = "sic")@changepoints))
```

## Hidden Markov models

A hidden Markov model is the discrete-state cousin of the state-space model: a
finite Markov chain over hidden states, each emitting a Gaussian observation.
`hmm()` builds the model, `hmm_filter()` gives the state posteriors, and
`hmm_viterbi()` returns the single most likely state path.

We build a two-state mean-shift model (a low state near `0` and a high state
near `4`) and simulate from it:

```{r hmm-build}
model <- hmm(
  init_prob = c(0.5, 0.5),
  transition = matrix(c(0.95, 0.05, 0.05, 0.95), nrow = 2, byrow = TRUE),
  emission_mean = c(0, 4), emission_sd = c(1, 1)
)
model

true_state <- c(rep(1L, 60), rep(2L, 60), rep(1L, 60))
y <- rnorm(180, mean = c(0, 4)[true_state], sd = 1)
```

The smoothed posterior and the Viterbi path both recover the hidden state with
near-perfect accuracy for these well-separated states:

```{r hmm-recover}
fit <- hmm_filter(model, y)
path <- hmm_viterbi(model, y)

c(smoothed_accuracy = mean(fit@state_path == true_state),
  viterbi_accuracy  = mean(path == true_state))
```

```{r hmm-plot, echo = FALSE, fig.cap = "Observations coloured by their true hidden state (grey = low, red = high), with the smoothed posterior probability of the high state overlaid."}
plot(y, col = c("grey70", "firebrick")[true_state], pch = 16, cex = 0.6,
     xlab = "time", ylab = "observation",
     main = "HMM: smoothed posterior of the high state")
par(new = TRUE)
plot(fit@smoothed[, 2], type = "l", col = "steelblue", lwd = 2,
     axes = FALSE, xlab = "", ylab = "", ylim = c(0, 1))
axis(4)
mtext("P(high state)", side = 4, line = 2)
```

The smoothed posterior (blue) rises to near one over the central high-state
block and falls back to near zero on either side. The difference between the
two read-outs is subtle but real: the Viterbi path is *jointly* optimal, the
single most likely sequence, while the smoothed mode is optimal at each step
taken on its own.

## References

Rabiner, L. R. (1989). A tutorial on hidden Markov models and selected
applications in speech recognition. *Proceedings of the IEEE*, 77(2),
257--286.

Scott, A. J. and Knott, M. (1974). A cluster analysis method for grouping
means in the analysis of variance. *Biometrics*, 30(3), 507--512. (The
binary-segmentation method `detect_changepoint()` implements.)

Killick, R., Fearnhead, P. and Eckley, I. A. (2012). Optimal detection of
changepoints with a linear computational cost. *Journal of the American
Statistical Association*, 107(500), 1590--1598. (See also, for the
linear-cost PELT alternative to the quadratic-cost binary segmentation used
here.)
