---
title: "The Elliott-van der Hoek-Malcolm spread model with kalmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Elliott-van der Hoek-Malcolm spread model with kalmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~18 s under rmarkdown::render(), dominated by ssm_fit()'s
     two-start optimiser; macOS arm64 (Darwin aarch64), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 150)
set.seed(2005)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kalmix)
```

## Why

An analyst pricing a mean-reversion trade on a grain basis, or on any other
spread, naturally fits an autoregression to what is actually observed: the
noisy spread series, not the true underlying spread that reverts. Elliott,
van der Hoek and Malcolm (2005) show why that is not innocuous: the spread a
trader sees is not the spread that mean-reverts, so fitting the observed
series directly overstates the speed of reversion and understates the
expected wait for a position to close. Their model founded the
mean-reverting-spread machinery in `kalmix`, on which *Pairs trading with
kalmix* is built, and this vignette reproduces it with the package's general
state-space tools, to check the same warning their paper makes: does
estimating through the filter, which separates the two noise sources,
actually recover the slower true dynamics where a naive fit on the raw
series does not?

## What

[`ssm()`](../reference/ssm.html) builds the paper's model directly, with a
state intercept for the mean-reverting AR(1) drift.
[`ssm_fit()`](../reference/ssm_fit.html) maximises the prediction-error
likelihood of the state-space form, separating the state noise from the
observation noise instead of lumping them together.
[`ou_fit()`](../reference/ou_fit.html) is the naive comparison: it fits an
autoregression directly to the observed series, with no separation of the
two noise sources. [`half_life()`](../reference/half_life.html) reads the
mean-reversion half-life off either fit.

## Do

### The 2005 model

The paper's state process (its eq. 1, rewritten as its eq. 7) is a Gaussian
AR(1) with an intercept,

$$x_{k+1} = A + B\,x_k + C\,\varepsilon_{k+1},
  \qquad \varepsilon_k \sim N(0, 1),$$

with $A = a\tau \geq 0$, $0 < B = 1 - b\tau < 1$ and $C = \sigma\sqrt{\tau}$,
mean-reverting to $\mu = a/b$. The trader observes the spread in Gaussian
noise (eq. 9),

$$y_k = x_k + D\,\omega_k, \qquad \omega_k \sim N(0, 1),$$

and the object of interest is the filter $\hat{x}_k = E[x_k \mid
y_0, \ldots, y_k]$ (eq. 10), which requires estimates of $(A, B, C, D)$.
With the state intercept of `ssm()` this is a four-number model
specification. The stationary moments seed the prior:

```{r model}
a_true <- 0.06
b_true <- 0.85
c_true <- 0.30
d_true <- 0.45
mu <- a_true / (1 - b_true)
var_x <- c_true^2 / (1 - b_true^2)

spread_2005 <- ssm(
  transition = b_true, observation = 1,
  state_cov = c_true^2, obs_cov = d_true^2,
  state_intercept = a_true,
  init_state = mu, init_cov = var_x
)
spread_2005
```

Simulating a synthetic series from the paper's equations gives a series
where the true hidden spread is known, so every later claim can be checked
against the truth:

```{r simulate}
n <- 600L
x <- numeric(n)
x[1] <- mu
for (k in 2:n) {
  x[k] <- a_true + b_true * x[k - 1] + c_true * rnorm(1)
}
y <- x + d_true * rnorm(n)
```

```{r simulate-plot, echo = has_ggplot2, eval = has_ggplot2, fig.cap = "The simulated observed spread (points) against the true hidden mean-reverting state (line) it was generated from."}
sim_data <- data.frame(k = seq_len(n), observed = y, hidden = x)

ggplot2::ggplot(sim_data, ggplot2::aes(x = k)) +
  ggplot2::geom_point(ggplot2::aes(y = observed), colour = "grey55",
                       size = 0.5) +
  ggplot2::geom_line(ggplot2::aes(y = hidden), colour = "#D55E00",
                      linewidth = 0.6) +
  ggplot2::labs(
    x = "k", y = "spread (synthetic units)",
    title = "Observed spread (points) and hidden state (line)"
  ) +
  ggplot2::theme_minimal()
```

```{r simulate-plot-skip, echo = FALSE, eval = !has_ggplot2, results = "asis"}
cat("ggplot2 is not installed on this build, so the simulated-spread",
    "figure is skipped.\n")
```

The grey points are what a trader actually sees; the orange line is the
hidden mean-reverting state the paper's model describes, known here only
because the series was simulated from it.

### Why least squares on the observed spread misprices the trade

Fitting the AR(1) directly to the noisy observations treats $y_k$ as if it
were $x_k$. Measurement error attenuates the estimated autoregression
towards zero, so the fitted mean-reversion looks faster than it is and the
implied half-life comes out too short: the naive trader expects the spread
to revert sooner than it does.

```{r naive}
naive <- ou_fit(y)
naive_half_life <- half_life(naive)
true_half_life <- log(2) / -log(b_true)
```

### Estimating the model through the filter

`ssm_fit()` maximises the prediction-error likelihood of the state-space
form, which separates the state noise $C$ from the observation noise $D$
instead of lumping them. A two-step estimator handles the intercept: the
sample mean estimates $\mu$ (the observation mean equals the state mean),
and $\hat{A} = \hat{\mu}(1 - \hat{B})$ recovers the intercept from the
fitted transition.

```{r fit}
mu_hat <- mean(y)
template <- ssm(
  transition = 0.5, observation = 1, state_cov = 1, obs_cov = 1,
  init_state = 0, init_cov = var(y)
)
fit <- ssm_fit(template, y - mu_hat, transition = TRUE, n_start = 2L)

b_hat <- fit@estimates$transition[1, 1]
c_hat <- sqrt(fit@estimates$state_cov[1, 1])
d_hat <- sqrt(fit@estimates$obs_cov[1, 1])
a_hat <- mu_hat * (1 - b_hat)
ss_half_life <- log(2) / -log(b_hat)
```

```{r fit-kable, echo = FALSE}
knitr::kable(
  rbind(
    truth = c(A = a_true, B = b_true, C = c_true, D = d_true),
    estimated = c(a_hat, b_hat, c_hat, d_hat)
  ),
  digits = 3,
  caption = paste(
    "The paper's four parameters: true simulation values against the",
    "state-space estimate."
  )
)
```

```{r half-life-kable, echo = FALSE}
knitr::kable(
  data.frame(
    method = c("true (from B)", "state-space (ssm_fit)", "naive (ou_fit)"),
    half_life = round(c(true_half_life, ss_half_life, naive_half_life), 3)
  ),
  col.names = c("Method", "Half-life (steps)"),
  caption = "The mean-reversion half-life implied by each estimate."
)
```

### The filter and the trading rule

The paper's trading rule (its eqs. 15 to 17) works on the
Ornstein-Uhlenbeck form with $\rho = b = (1 - B)/\tau$ and
$\mu = a/b$: enter a pair trade when the observed spread leaves the band
$\mu \pm c\,\sigma/\sqrt{2\rho}$ and unwind when the state is expected back
at $\mu$. The filtered state is the quantity those decisions condition on:

```{r filter}
fitted_model <- ssm(
  transition = b_hat, observation = 1,
  state_cov = c_hat^2, obs_cov = d_hat^2,
  state_intercept = a_hat,
  init_state = mu_hat, init_cov = c_hat^2 / (1 - b_hat^2)
)
kf <- kalman_filter(fitted_model, y)

tau <- 1
rho <- (1 - b_hat) / tau
sigma <- c_hat / sqrt(tau)
band <- 1 * sigma / sqrt(2 * rho)
```

```{r filter-plot, echo = has_ggplot2, eval = has_ggplot2, fig.cap = "The observed spread (points) with the Kalman-filtered estimate (line, with its 95 per cent band) and the entry band implied by the paper's trading rule (dashed)."}
filter_se <- sqrt(vapply(kf@filtered_cov, function(p) p[1, 1], numeric(1)))
filter_data <- data.frame(
  k = seq_len(n), observed = y, filtered = kf@filtered_mean[, 1],
  lower = kf@filtered_mean[, 1] - 1.96 * filter_se,
  upper = kf@filtered_mean[, 1] + 1.96 * filter_se
)
band_lines <- data.frame(y = mu_hat + c(-band, 0, band))

ggplot2::ggplot(filter_data, ggplot2::aes(x = k)) +
  ggplot2::geom_point(ggplot2::aes(y = observed), colour = "grey55",
                       size = 0.5) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lower, ymax = upper),
                        fill = "#0072B2", alpha = 0.2) +
  ggplot2::geom_line(ggplot2::aes(y = filtered), colour = "#0072B2",
                      linewidth = 0.6) +
  ggplot2::geom_hline(data = band_lines, ggplot2::aes(yintercept = y),
                       linetype = "dashed", colour = "black") +
  ggplot2::labs(
    x = "k", y = "spread (synthetic units)",
    title = "Filtered spread with the entry band"
  ) +
  ggplot2::theme_minimal()
```

```{r filter-plot-skip, echo = FALSE, eval = !has_ggplot2, results = "asis"}
cat("ggplot2 is not installed on this build, so the filtered-spread",
    "figure is skipped.\n")
```

### The single-regime base case of the regime filter

`kalmix_filter()` generalises this model to volatility regimes; with
`n_regimes = 1` it reduces to exactly the 2005 filter, which makes the
paper's model the base case of the package's regime machinery:

```{r base-case}
one_regime <- kalmix_filter(naive, y, n_regimes = 1L)
one_regime_class <- class(one_regime)[1]
```

## Read

The naive fit's half-life of `r round(naive_half_life, 3)` steps
understates the true value of `r round(true_half_life, 3)` steps by
`r round(100 * (true_half_life - naive_half_life) / true_half_life, 1)`
per cent, exactly the direction the paper warns of: measurement error
attenuates the naive autoregression towards faster reversion. The
state-space estimate, at `r round(ss_half_life, 3)` steps, sits far closer
to the truth, because `ssm_fit()` separates the state noise `C` from the
observation noise `D` where the naive fit conflates them, visible directly
in the parameter table: the naive route has no way to represent `D` at all,
while the state-space route recovers `A`, `B`, `C` and `D` all close to
their simulated values. The first figure shows what the naive fit is misled
by: the grey observation cloud around the orange truth is wide relative to
how tightly the truth itself reverts, so a fit that cannot see the two
noise sources separately reads part of that observation scatter as fast
mean-reversion. The second figure shows the corrected picture: the blue
filtered line and its band track the orange-obscured truth far more
closely than the raw points do, and the dashed entry band marks where the
paper's trading rule would act.

`kalmix_filter(naive, y, n_regimes = 1L)` returns an object of class
`r one_regime_class`, the same class `kalman_filter()` itself returns:
with one regime the general regime machinery is not merely similar to the
2005 filter, it collapses to it exactly, confirming the paper's model is the
base case of the package's regime-switching generalisation rather than a
separate implementation.

## Limits

The comparison here isolates one effect, observation noise attenuating a
naive autoregressive fit, on data simulated from the paper's own equations,
so the state-space route is checked against a truth it was built to recover
under exactly the conditions it assumes. A real spread need not follow this
Gaussian AR(1) exactly, and no adequacy check is run on either fit here to
ask whether it does; that machinery, and the abstention it produces when a
model fails it, is shown in *Assimilative causal inference with kalmix*.
`ssm_fit()`'s standard errors, not used in this vignette, are asymptotic and
can be optimistic on a series this short, where the likelihood separates the
two noise sources only weakly; a wide interval there would be a finding
about the data, not a defect of the fit.

## What to read next

*Pairs trading with kalmix* applies this same mean-reverting-spread
machinery, through `ou_fit()` and `kalmix_filter()`, to a volatility-regime
question this vignette does not address. *Getting started: state-space
models with kalmix* covers `ssm()`, `kalman_filter()` and the general
recursion this vignette specialises to the paper's model. *Assimilative
causal inference with kalmix* covers the model-adequacy grounding neither
fit shown here is checked against.

## Reproduce

Seed `2005`, chosen for the paper's publication year, for the simulation.

```{r session-info}
sessionInfo()
```
