---
title: "The Elliott-van der Hoek-Malcolm spread model with kalmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Elliott-van der Hoek-Malcolm spread model with kalmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
set.seed(2005)
library(kalmix)
```

Elliott, van der Hoek and Malcolm (2005) model a pairs-trading spread as a
mean-reverting Gaussian state observed in noise, and they are explicit that
the observation noise matters: the spread the trader sees is not the spread
that mean-reverts. This vignette reproduces their model with the general
`kalmix` machinery and demonstrates the point the state-space treatment
exists to make. Estimating the dynamics directly from the noisy spread by
ordinary least squares misprices the trade, while the filter-based
estimation recovers the true dynamics.

## The 2005 model

The paper's state process (its eq. 1, rewritten as its eq. 7) is a Gaussian
AR(1) with an intercept,

$$x_{k+1} = A + B\,x_k + C\,\varepsilon_{k+1},
  \qquad \varepsilon_k \sim N(0, 1),$$

with $A = a\tau \ge 0$, $0 < B = 1 - b\tau < 1$ and $C = \sigma\sqrt{\tau}$,
mean-reverting to $\mu = a/b$. The trader observes the spread in Gaussian
noise (eq. 9),

$$y_k = x_k + D\,\omega_k, \qquad \omega_k \sim N(0, 1),$$

and the object of interest is the filter $\hat{x}_k = E[x_k \mid
y_0, \ldots, y_k]$ (eq. 10), which requires estimates of $(A, B, C, D)$.

With the state intercept of `ssm()` this is a four-number model
specification. The stationary moments seed the prior:

```{r model}
A <- 0.06
B <- 0.85
C <- 0.30
D <- 0.45
mu <- A / (1 - B)
var_x <- C^2 / (1 - B^2)

spread_2005 <- ssm(
  transition = B, observation = 1,
  state_cov = C^2, obs_cov = D^2,
  state_intercept = A,
  init_state = mu, init_cov = var_x
)
spread_2005
```

Simulating from the paper's equations gives a series where the true hidden
spread is known, so every later claim can be checked against the truth:

```{r simulate, fig.cap = "The simulated observed spread (grey points) against the true hidden mean-reverting state (red line) it was generated from."}
n <- 600
x <- numeric(n)
x[1] <- mu
for (k in 2:n) {
  x[k] <- A + B * x[k - 1] + C * rnorm(1)
}
y <- x + D * rnorm(n)

plot(y, type = "p", cex = 0.4, col = "grey50",
     xlab = "k", ylab = "spread",
     main = "Observed spread (points) and hidden state (line)")
lines(x, col = "firebrick")
```

The grey points are what a trader actually sees; the red line is the hidden
mean-reverting state the paper's model describes, known here only because
the series was simulated from it.

## Why least squares on the observed spread misprices the trade

Fitting the AR(1) directly to the noisy observations treats $y_k$ as if it
were $x_k$. Measurement error attenuates the estimated autoregression
toward zero, so the fitted mean-reversion looks faster than it is and the
implied half-life comes out too short: the naive trader expects the spread
to revert sooner than it does.

```{r naive}
naive <- ou_fit(y)
half_life(naive)

## The true half-life implied by B.
log(2) / -log(B)
```

## Estimating the model through the filter

`ssm_fit()` maximises the prediction-error likelihood of the state-space
form, which separates the state noise $C$ from the observation noise $D$
instead of lumping them. A two-step estimator handles the intercept: the
sample mean estimates $\mu$ (the observation mean equals the state mean),
and $\hat{A} = \hat{\mu}(1 - \hat{B})$ recovers the intercept from the
fitted transition.

```{r fit}
mu_hat <- mean(y)
template <- ssm(
  transition = 0.5, observation = 1, state_cov = 1, obs_cov = 1,
  init_state = 0, init_cov = var(y)
)
fit <- ssm_fit(template, y - mu_hat, transition = TRUE, n_start = 2L)

B_hat <- fit@estimates$transition[1, 1]
C_hat <- sqrt(fit@estimates$state_cov[1, 1])
D_hat <- sqrt(fit@estimates$obs_cov[1, 1])
A_hat <- mu_hat * (1 - B_hat)

round(rbind(
  truth     = c(A = A, B = B, C = C, D = D),
  estimated = c(A_hat, B_hat, C_hat, D_hat)
), 3)

## The state-space half-life against the naive one.
c(state_space = log(2) / -log(B_hat), naive = unname(half_life(naive)))
```

The filter-based estimate recovers the slow true dynamics where the naive
fit reports a half-life that is too short. In trading terms the naive fit
would enter and unwind too eagerly, and the paper's framework exists
precisely to avoid that.

## The filter and the trading rule

The paper's trading rule (its eqs. 15 to 17) works on the
Ornstein-Uhlenbeck form with $\rho = b = (1 - B)/\tau$ and
$\mu = a/b$: enter a pair trade when the observed spread leaves the band
$\mu \pm c\,\sigma/\sqrt{2\rho}$ and unwind when the state is expected back
at $\mu$. The filtered state is the quantity those decisions condition on:

```{r filter, fig.cap = "The observed spread (grey points) with the Kalman-filtered estimate (blue line) and the entry band (dashed) implied by the paper's trading rule."}
fitted_model <- ssm(
  transition = B_hat, observation = 1,
  state_cov = C_hat^2, obs_cov = D_hat^2,
  state_intercept = A_hat,
  init_state = mu_hat, init_cov = C_hat^2 / (1 - B_hat^2)
)
kf <- kalman_filter(fitted_model, y)

tau <- 1
rho <- (1 - B_hat) / tau
sigma <- C_hat / sqrt(tau)
band <- 1 * sigma / sqrt(2 * rho)

plot(y, type = "p", cex = 0.4, col = "grey50",
     xlab = "k", ylab = "spread",
     main = "Filtered spread with the entry band")
lines(kf@filtered_mean[, 1], col = "steelblue")
abline(h = mu_hat + c(-band, 0, band), lty = c(2, 1, 2))
```

The filtered path (blue) strips the observation noise the naive approach
mistakes for mean-reversion, and the band marks the entry threshold at
$c = 1$ standard deviations of the stationary state.

## The single-regime base case of the regime filter

`kalmix_filter()` generalises this model to volatility regimes; with
`n_regimes = 1` it reduces to exactly the 2005 filter, which makes the
paper's model the base case of the package's regime machinery:

```{r base-case}
model <- ou_fit(y)
one_regime <- kalmix_filter(model, y, n_regimes = 1L)
class(one_regime)
```

The adequacy battery of `innovation_diagnostics()` then certifies the
fitted specification the same way it certifies every state-space model in
the package, and the assimilative-causal-inference read-out of `aci()`
applies unchanged. The spread model that founded the package is one
`ssm()` call within it.

## References

Elliott, R. J., van der Hoek, J. and Malcolm, W. P. (2005). Pairs trading.
*Quantitative Finance*, 5(3), 271--276. doi:10.1080/14697680500149370.
