---
title: "Pairs trading with kalmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pairs trading with kalmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS arm64 (Darwin
     aarch64), R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 150)
set.seed(1)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kalmix)
```

## Why

A grain marketer watching the gap between the local delivered price and a
reference benchmark asks a mean-reversion question: is this week's wider
spread a genuine opportunity that will close back to normal on its own, or
just as likely to widen further, and how volatile is the spread right now
compared with a calmer stretch of the year? The same question recurs
wherever two series share a common driver but drift apart temporarily, and
it is exactly the question the classical pairs-trading problem asks of two
market prices, which is the application `kalmix`'s mean-reverting-spread
machinery was first built for, on the scalar Ornstein-Uhlenbeck filter of
Elliott, van der Hoek and Malcolm (2005). This vignette works the recipe on
a synthetic pair of markets: it builds a spread, fits its mean reversion,
filters its volatility regime with the native mixture filter, and reads a
causal contrast off a simulated event. The identical construction runs
unchanged on a grain basis, or on any other pair of cointegrated series.

## What

[`spread_series()`](../reference/spread_series.html) builds the stationary
residual of a cointegration regression between two series.
[`ou_fit()`](../reference/ou_fit.html) estimates the continuous-time
Ornstein-Uhlenbeck dynamics of a spread, returning a
[`spread_model`](../reference/spread_model.html) with a `kappa`
mean-reversion speed, `theta` long-run mean, `sigma` volatility, and
[`half_life()`](../reference/half_life.html). `kalmix_filter()` expands a
fitted spread into volatility regimes and filters it through the general
mixture engine. `its_causal()` forecasts a counterfactual from pre-event
dynamics and contrasts it with what was observed.

## Do

### A cointegration spread

We simulate two cointegrated synthetic markets, then recover the spread
between them: market `b` is a random walk, and market `a` tracks it with a
hedge ratio of two plus a mean-reverting residual `u` (a stationary AR(1)
process, the realistic cointegration case).

```{r spread}
b <- cumsum(rnorm(500))
u <- as.numeric(arima.sim(list(ar = 0.8), n = 500))
a <- 2 * b + u

spread <- spread_series(a, b)
hedge_ratio <- attr(spread, "hedge_ratio")
```

The hedge ratio is recovered as `r round(hedge_ratio, 3)`, close to its true
value of two, and `spread` is the stationary residual a pairs trade is
built on.

### Fitting the mean reversion

`ou_fit()` estimates the mean-reversion speed `kappa`, the long-run mean
`theta`, the volatility `sigma`, and the implied half-life, from the
continuous-time model

$$dX_t = \kappa (\theta - X_t)\, dt + \sigma\, dW_t,$$

so the half-life of a deviation is $\log 2 / \kappa$.

```{r ou}
model <- ou_fit(spread)
model

spread_half_life <- half_life(model)
```

The half-life is the expected time for a deviation from the long-run mean to
halve, the single most useful number for sizing a mean-reversion position: a
half-life of `r round(spread_half_life, 2)` steps here.

A series whose dynamics are inconsistent with an Ornstein-Uhlenbeck process
(here an anti-persistent series with negative autocorrelation) is refused
rather than fitted with a meaningless half-life:

```{r refuse, error = TRUE}
ou_fit(as.numeric(arima.sim(list(ar = -0.6), n = 500)))
```

### Tracking the volatility regime

A spread is not equally tradeable at all times: it passes through calm
periods and turbulent periods, the way a grain basis is typically calmer
through much of the marketing year and more volatile around harvest, when
local delivery pressure peaks. `kalmix_filter()` expands the fitted spread
into volatility regimes (linear-Gaussian state-space models that share the
spread dynamics but differ in observation noise) and filters the spread
through the general `mixture_filter()` engine, needing no other package. We
build a spread that is calm in its first half and turbulent in its second:

```{r regime}
calm <- as.numeric(arima.sim(list(ar = 0.7), n = 200, sd = 0.4))
wild <- as.numeric(arima.sim(list(ar = 0.7), n = 200, sd = 2))
two_regime <- c(calm, wild)

regime_model <- ou_fit(two_regime)
regime_fitted <- kalmix_filter(regime_model, two_regime, n_regimes = 2)
calm_prob <- mean(regime_fitted@regime_prob[1:200, 2])
wild_prob <- mean(regime_fitted@regime_prob[201:400, 2])
```

```{r regime-kable, echo = FALSE}
knitr::kable(
  data.frame(
    half = c("calm (steps 1 to 200)", "turbulent (steps 201 to 400)"),
    mean_prob = round(c(calm_prob, wild_prob), 3)
  ),
  col.names = c("Half of the series", "Mean P(turbulent regime)"),
  caption = paste(
    "Filtered probability of the turbulent regime, averaged over each",
    "half."
  )
)
```

```{r regime-plot, echo = has_ggplot2, eval = has_ggplot2, fig.cap = "The spread series (points) with the filtered probability of the turbulent regime (line, right-hand axis)."}
range_y <- range(two_regime)
span_y <- diff(range_y)
prob_turbulent <- regime_fitted@regime_prob[, 2]
prob_scaled <- prob_turbulent * span_y + min(range_y)

regime_data <- data.frame(
  step = seq_along(two_regime), spread = two_regime,
  half = factor(rep(c("calm", "turbulent"), each = 200),
                levels = c("calm", "turbulent")),
  posterior = prob_scaled
)

ggplot2::ggplot(regime_data, ggplot2::aes(x = step)) +
  ggplot2::geom_point(ggplot2::aes(y = spread, colour = half), size = 0.8) +
  ggplot2::geom_line(ggplot2::aes(y = posterior), colour = "black",
                      linewidth = 0.6) +
  ggplot2::scale_colour_viridis_d(name = "true half", end = 0.8) +
  ggplot2::scale_y_continuous(
    name = "spread (synthetic units)",
    sec.axis = ggplot2::sec_axis(
      ~ (. - min(range_y)) / span_y, name = "P(turbulent regime)"
    )
  ) +
  ggplot2::labs(x = "step",
                title = "Filtered turbulent-regime probability") +
  ggplot2::theme_minimal()
```

```{r regime-plot-skip, echo = FALSE, eval = !has_ggplot2, results = "asis"}
cat("ggplot2 is not installed on this build, so the regime-probability",
    "figure is skipped.\n")
```

### A causal read on an event

Suppose a structural event, a rail-freight surcharge or an export-parity
shift, is suspected to have moved the spread at a known time. `its_causal()`
forecasts the counterfactual spread the pre-event dynamics imply and
contrasts it with what was observed; the interval separates the two
outcomes, a real effect excludes zero and a non-effect straddles it.

```{r its}
pre <- as.numeric(arima.sim(list(ar = 0.8), n = 80, sd = 0.5))
post <- as.numeric(arima.sim(list(ar = 0.8), n = 40, sd = 0.5)) + 2
impact <- its_causal(c(pre, post), intervention = 81L)
impact

interval_spans_zero <- impact@avg_effect_lower <= 0 &&
  impact@avg_effect_upper >= 0
```

## Read

The hedge ratio recovered from the cointegration regression,
`r round(hedge_ratio, 3)`, sits close to the true value of two used to
simulate the pair, and the fitted Ornstein-Uhlenbeck half-life of
`r round(spread_half_life, 2)` steps is the number a mean-reversion position
would be sized against. The regime table and figure agree with each other
and with how the two halves of the series were built: the turbulent-regime
probability averages `r round(wild_prob, 3)` over the turbulent half against
`r round(calm_prob, 3)` over the calm half, and the black posterior line in
the figure rises sharply at the step-200 boundary, tracking the visibly
wider scatter of the turbulent points rather than lagging behind it.

The event contrast recovers an average effect of `r round(impact@avg_effect, 3)`
against the two-unit lift injected into the post-period, with a
`r round(100 * impact@level)` per cent interval of
[`r round(impact@avg_effect_lower, 3)`, `r round(impact@avg_effect_upper, 3)`]
that `r if (interval_spans_zero) "straddles" else "excludes"` zero, so the
simulated event is correctly flagged as a genuine shift rather than noise.

The pairs-trading workflow here is built entirely from the general
machinery: `kalmix_filter()` is a thin wrapper over `mixture_filter()`,
itself the regime-switching generalisation of `kalman_filter()`, and
`its_causal()` forecasts with the same `ssm()` and Kalman recursion covered
in *Getting started: state-space models with kalmix*. The mean-reverting
spread is one application of tools that serve any latent state observed
through noise, on a grain basis exactly as on the synthetic pair worked
through here.

## Limits

The two markets simulated here are synthetic, chosen to isolate the
mean-reversion and regime-switching mechanics from any question about
whether real prices are actually cointegrated, which this vignette does not
test. `ou_fit()` estimates the spread's dynamics directly from the observed
series, folding observation noise into the fitted volatility rather than
separating the two; *The Elliott-van der Hoek-Malcolm spread model with
kalmix* shows why that distinction matters and how the general `ssm_fit()`
machinery separates them. The event contrast from `its_causal()` assumes the
pre-event dynamics would have continued unchanged absent the event, an
assumption this vignette does not defend beyond simulating a series built to
satisfy it. No model-adequacy check is run on any model fitted here; a
spread whose dynamics are inconsistent with an Ornstein-Uhlenbeck process is
refused outright by `ou_fit()`, as shown above, but a model that is
technically fittable is not thereby checked for adequacy the way `aci()`'s
read-out is in *Assimilative causal inference with kalmix*.

## What to read next

*The Elliott-van der Hoek-Malcolm spread model with kalmix* reproduces the
founding paper's model with the general `ssm()` machinery and shows why
separating observation noise from state noise changes the estimated
half-life. *Getting started: state-space models with kalmix* covers the
`kalman_filter()` and `rts_smoother()` this vignette's regime filter and
event contrast are both built from. *Assimilative causal inference with
kalmix* grounds a causal read-out against a model-adequacy check, including
the abstention it produces on a genuinely inadequate model.

## Reproduce

Seed `1` for every simulation in this vignette.

```{r session-info}
sessionInfo()
```
