---
title: "Pairs trading with kalmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pairs trading with kalmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
set.seed(1)
library(kalmix)
```

Mean-reverting spread estimation for pairs trading is one application of the
general state-space machinery in `kalmix`. A pairs trade bets that the spread
between two cointegrated assets is stationary: it reverts to a long-run mean,
so an unusually wide spread is an opportunity. This vignette builds a spread,
fits its mean reversion, filters its volatility regime with the native mixture
filter, and reads a causal contrast off an event with `its_causal()`.

## A cointegration spread

We simulate two cointegrated series -- `b` is a random walk, and `a` tracks it
with a hedge ratio of two plus a mean-reverting spread `u` (a stationary AR(1)
process, the realistic cointegration case) -- then recover the spread:

```{r spread}
b <- cumsum(rnorm(500))
u <- as.numeric(arima.sim(list(ar = 0.8), n = 500))
a <- 2 * b + u

spread <- spread_series(a, b)
attr(spread, "hedge_ratio")
```

The hedge ratio is recovered close to its true value of two, and `spread` is
the stationary residual a pairs trade is built on.

## Fitting the mean reversion

`ou_fit()` estimates the continuous-time Ornstein-Uhlenbeck parameters of the
spread: the mean-reversion speed `kappa`, the long-run mean `theta`, the
volatility `sigma`, and the implied half-life. The continuous-time model is

$$dX_t = \kappa (\theta - X_t)\, dt + \sigma\, dW_t,$$

so the half-life of a deviation is $\log 2 / \kappa$.

```{r ou}
model <- ou_fit(spread)
model

half_life(model)
```

The half-life is the expected time for a deviation from the long-run mean to
halve -- the single most useful number for sizing a mean-reversion trade.

A series whose dynamics are inconsistent with an Ornstein-Uhlenbeck process --
here an anti-persistent series with negative autocorrelation -- is refused
rather than fitted with a meaningless half-life:

```{r refuse, error = TRUE}
ou_fit(as.numeric(arima.sim(list(ar = -0.6), n = 500)))
```

## Tracking the volatility regime

A spread is not equally tradeable at all times: it passes through calm periods
and turbulent periods. `kalmix_filter()` expands the fitted spread into
volatility regimes -- linear-Gaussian state-space models that share the spread
dynamics but differ in observation noise -- and filters the spread through the
general `mixture_filter()` engine. The result is the filtered probability of
each regime at every step.

This runs on the package's own native filter and needs no other package. We
build a spread that is calm in its first half and turbulent in its second, and
the filter raises the turbulent-regime probability accordingly:

```{r regime}
calm <- as.numeric(arima.sim(list(ar = 0.7), n = 200, sd = 0.4))
wild <- as.numeric(arima.sim(list(ar = 0.7), n = 200, sd = 2))
two_regime <- c(calm, wild)

fit <- kalmix_filter(ou_fit(two_regime), two_regime, n_regimes = 2)
c(calm_half = mean(fit@regime_prob[1:200, 2]),
  wild_half = mean(fit@regime_prob[201:400, 2]))
```

```{r regime-plot, echo = FALSE}
plot(two_regime, col = "grey70", pch = 16, cex = 0.5, xlab = "time",
     ylab = "spread", main = "Filtered turbulent-regime probability")
par(new = TRUE)
plot(fit@regime_prob[, 2], type = "l", col = "firebrick", lwd = 2,
     axes = FALSE, xlab = "", ylab = "", ylim = c(0, 1))
axis(4)
mtext("P(turbulent)", side = 4, line = 2)
```

The turbulent-regime probability (red) is low through the calm first half and
high through the turbulent second half -- the signal a mean-reversion strategy
would condition position size on.

## A causal read on an event

Suppose a structural event -- a corporate action, a regime change -- is
suspected to have shifted the spread at a known time. `its_causal()` forecasts
the counterfactual spread the pre-event dynamics imply, and contrasts it with
what was observed. Its interval is honest: a real effect excludes zero, a
non-effect straddles it.

```{r its}
pre <- as.numeric(arima.sim(list(ar = 0.8), n = 80, sd = 0.5))
post <- as.numeric(arima.sim(list(ar = 0.8), n = 40, sd = 0.5)) + 2
impact <- its_causal(c(pre, post), intervention = 81L)
impact
```

The estimated average effect recovers the injected lift and its interval
excludes zero, so the event is flagged as a genuine shift rather than noise.

## How kalmix fits together

The pairs-trading workflow here is built entirely from the general machinery:
`kalmix_filter()` is a thin wrapper over `mixture_filter()`, which is itself
the regime-switching generalisation of the `kalman_filter()`, and `its_causal()`
forecasts with the same `ssm()` and Kalman recursion. The mean-reverting spread
is one application; the same tools serve any latent state observed through
noise. The optional `engine = "proxymix"` path lets the regime filter defer to
a sibling package's mixture-Kalman engine where one is installed, but the native
filter is the default and always available.

## References

Elliott, R. J., van der Hoek, J. and Malcolm, W. P. (2005). Pairs trading.
*Quantitative Finance*, 5(3), 271--276.

Brodersen, K. H., Gallusser, F., Koehler, J., Remy, N. and Scott, S. L. (2015).
Inferring causal impact using Bayesian structural time-series models. *Annals
of Applied Statistics*, 9(1), 247--274.
