---
title: "Getting started with kalmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with kalmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`kalmix` models the mean-reverting spread that underlies a pairs trade. The
standalone core does two things: it builds a cointegration spread from two
price series, and it fits the spread's Ornstein-Uhlenbeck dynamics. The
regime-switching mixture Kalman filter is described in the *kalmix roadmap*
vignette.

## A cointegration spread

We simulate two cointegrated series -- `b` is a random walk, and `a` tracks
it with a hedge ratio of two plus a mean-reverting spread `u` (a stationary
AR(1) process, the realistic cointegration case) -- then recover the spread.

```{r spread}
library(kalmix)

set.seed(1)
b <- cumsum(rnorm(500))
u <- as.numeric(arima.sim(list(ar = 0.8), n = 500))
a <- 2 * b + u

spread <- spread_series(a, b)
attr(spread, "hedge_ratio")
```

The hedge ratio is recovered close to its true value of two, and `spread` is
the stationary residual a pairs trade is built on.

## Fitting the mean reversion

`ou_fit()` estimates the continuous-time Ornstein-Uhlenbeck parameters of the
spread by autoregression: the mean-reversion speed `kappa`, the long-run mean
`theta`, the volatility `sigma`, and the implied half-life.

```{r ou}
model <- ou_fit(spread)
model

half_life(model)
```

The half-life is the expected time for a deviation from the long-run mean to
halve -- the single most useful number for sizing a mean-reversion trade.

A series whose dynamics are inconsistent with an Ornstein-Uhlenbeck process
-- here an anti-persistent series with negative autocorrelation -- is refused
rather than fitted with a meaningless half-life:

```{r refuse, error = TRUE}
ou_fit(as.numeric(arima.sim(list(ar = -0.6), n = 500)))
```
