---
title: "The kalmix roadmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The kalmix roadmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`kalmix` is the applied state-space and statistical-arbitrage member of the
research team. Its headline capability -- filtering a regime-switching spread
state -- is supplied by a primitive it has commissioned from `proxymix`, and
this vignette records what is shipped, what is dormant, and how `kalmix`
connects to the rest of the team.

## What is shipped, what is dormant

The standalone core (`spread_series()`, `ou_fit()`, the `spread_model` class)
works on its own, with no sibling package installed. It is the scalar
mean-reverting spread of Elliott, van der Hoek and Malcolm (2005): a
cointegration spread whose Ornstein-Uhlenbeck dynamics are estimated directly.

The headline `kalmix_filter()` is **dormant-but-correct**. It filters the
latent spread state through a finite Gaussian-mixture Kalman recursion, where
each mixture component is one cointegration regime. That recursion is the
regime-switching generalisation of the 2005 scalar filter, and it is supplied
by `proxymix::gmm_filter()` -- a primitive that `kalmix` has commissioned but
that has not yet shipped in `proxymix`. Until it does, `kalmix_filter()`
validates its arguments and then errors with a commissioning notice:

```{r dormant, error = TRUE}
library(kalmix)
m <- ou_fit(as.numeric(arima.sim(list(ar = 0.85), n = 400)))
kalmix_filter(m, m@theta + rnorm(50))
```

The signature is stable, so analytics and tests can be written against it
today; activation is a single deliberate edit once the engine lands.

## How kalmix connects to the team

| Connects to | Practical contribution |
|---|---|
| `proxymix` (engine) | Commissions and reviews the mixture-Kalman primitive `gmm_filter()`; the scalar 2005 spread is the single-component base case, multi-regime spreads need the mixture. |
| `flexyBayes` (leader) | Contributes a real econometric model class -- the spread state-space model -- to the leader's inference and triangulation mandate. |
| `kernR` (validation) | Cointegration and regime-break testing, and the causal-direction question of which leg leads the other. |
| MaxAIbase (brain) | The project's decision ledger is ingested at the commercial tier; `kalmix` serves as the team's applied quantitative-finance reviewer. |

## Membership posture

`kalmix` is a knowledge-sharing and constructive-reviewer member that is also
dormant-wired to `proxymix`. It is standalone-functional today and ready to
activate its sibling adapters with a one-line edit when the primitives ship --
the readiness the team asks of every member.

## Reference

Elliott, R. J., van der Hoek, J. and Malcolm, W. P. (2005). Pairs trading.
*Quantitative Finance*, 5(3), 271--276.
