---
title: "Getting started: state-space models with kalmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: state-space models with kalmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
set.seed(1)
```

`kalmix` is a compact toolkit for state-space models. A state-space model
describes a latent state that evolves over time and is seen only through noisy
observations; the recursions in this package recover the state, smooth it given
the whole series, detect regime switches and change-points, and read a causal
contrast off a forecast. This vignette covers the linear-Gaussian core --
the `ssm()` model, the `kalman_filter()`, and the `rts_smoother()`. The
*Change-points and hidden states with kalmix* vignette covers discrete-state
models, and *Pairs trading with kalmix* covers the mean-reverting-spread
application.

## A linear-Gaussian state-space model

The general model has a latent state $x_t$ that evolves through a transition
matrix and Gaussian process noise, observed through an observation matrix and
Gaussian observation noise:

$$x_t = A_t\, x_{t-1} + w_t, \qquad w_t \sim N(0, Q_t),$$
$$y_t = B_t\, x_t + v_t, \qquad v_t \sim N(0, R_t).$$

`ssm()` builds this from the four system matrices plus a prior on the initial
state. The simplest case is the local level -- a random walk seen through
observation noise -- which can be written with bare scalars:

```{r local-level}
library(kalmix)

model <- ssm(
  transition = 1, observation = 1,
  state_cov = 0.01, obs_cov = 1,
  init_state = 0, init_cov = 10
)
model
```

## Filtering: recovering the state online

`kalman_filter()` runs the forward recursion of Kalman (1960). At each step it
predicts the state forward and then corrects it with the new observation. We
simulate a slowly-drifting level seen through noise, then recover it:

```{r filter}
level <- cumsum(rnorm(150, sd = 0.1))
y <- level + rnorm(150)

fit <- kalman_filter(model, y)
fit
```

The filtered mean tracks the latent level far more closely than the raw
observations do:

```{r filter-plot, echo = FALSE}
plot(y, col = "grey70", pch = 16, cex = 0.6, xlab = "time", ylab = "value",
     main = "Filtered state vs noisy observations")
lines(level, col = "black", lwd = 2)
lines(fit@filtered_mean[, 1], col = "steelblue", lwd = 2)
legend("topleft", c("truth", "observed", "filtered"),
       col = c("black", "grey70", "steelblue"),
       lwd = c(2, NA, 2), pch = c(NA, 16, NA), bty = "n")
```

## Smoothing: using the whole series

The filter conditions only on the past. The Rauch-Tung-Striebel smoother
(Rauch, Tung and Striebel, 1965) runs a backward pass to condition every state
estimate on the whole series, which is strictly more informative:

```{r smooth}
smoothed <- rts_smoother(fit)

rmse <- function(a, b) sqrt(mean((a - b)^2))
c(filtered = rmse(fit@filtered_mean[, 1], level),
  smoothed = rmse(smoothed@smoothed_mean[, 1], level))
```

The smoothed estimate has the lower error, as it must -- it has seen the
future as well as the past.

## A vector state

The state can have more than one dimension. A constant-velocity model tracks a
position and an unobserved velocity; only the position is measured. The model
recovers the hidden velocity from the position series alone:

```{r velocity}
velocity_model <- ssm(
  transition = matrix(c(1, 0, 1, 1), nrow = 2),
  observation = matrix(c(1, 0), nrow = 1),
  state_cov = diag(c(1e-5, 1e-5)),
  obs_cov = 1,
  init_state = c(0, 1),
  init_cov = diag(c(10, 10))
)

truth_velocity <- 0.5
position <- truth_velocity * seq_len(150) + rnorm(150)
vfit <- kalman_filter(velocity_model, position)

## recovered terminal velocity, against the true 0.5
vfit@filtered_mean[150, 2]
```

The recovered velocity sits close to the true `0.5`, even though velocity was
never observed -- the model infers it from the rate of change of the position.

## References

Kalman, R. E. (1960). A new approach to linear filtering and prediction
problems. *Journal of Basic Engineering*, 82(1), 35--45.

Rauch, H. E., Tung, F. and Striebel, C. T. (1965). Maximum likelihood estimates
of linear dynamic systems. *AIAA Journal*, 3(8), 1445--1450.

Durbin, J. and Koopman, S. J. (2012). *Time Series Analysis by State Space
Methods*, 2nd ed. Oxford University Press.
